A fast spectral method for the inelastic Boltzmann collision operator and application to heated granular gases\tnoteref{ack}

### 2018-09-10 paper/paper_rev/fast_spec_inelastic_rev.tex

\documentclass[review, times]{elsarticle}

@@

\usepackage{color}
\modulolinenumbers[5]

@@

Now assuming $f$ is a solution to equation \eqref{inBoltz}, the conservation of mass and momentum is straightforward by taking $\varphi = 1$ and $v$ respectively in the weak form \eqref{weak22}, i.e.,
\begin{equation}
\rho \equiv \rho_0, \quad  m\equiv m_0,
\end{equation}
with $\rho_0$ and $m_0$ being the initial density and momentum. By taking $\varphi=|v|^2/2$ in \eqref{weak22}, one obtains the following energy equation
\begin{equation}\label{ki_energy}
\partial_t E -\varepsilon d\rho_0=-\frac{1}{16}\int_{\mathbb{R}^d}\int_{\mathbb{R}^d}\int_{S^{d-1}} (1-e^2)B_{\sigma}(|g|,\sigma\cdot \hat{g})|g|^2(1-\sigma\cdot \hat{g})ff_*\,\rd{\sigma}\rd{v}\rd{v_*},
\end{equation}
where we used (\ref{momentum}) and (\ref{relation}).

@@

\subsubsection{3D inelastic Maxwell molecules}

@@

\begin{table}[htp!]
  \centering
  \begin{tabu} to 0.8\linewidth {X[1,c] X[2,c]}
    \toprule
    $M_{\text{sph}}$ & $|T_\text{num} - T_\text{ref}|$\\
    \midrule
    6 & 0.0013957739849754791 \\
    12 & 9.9706271716293315e-05 \\
    32 & 2.2499901350947482e-06 \\
    48 & 2.4272557155313734e-06 \\
    70 & 2.4703481364962698e-06 \\
    94 & 2.4703481364962698e-06 \\
    120 & 2.453380453903975e-06 \\
    \bottomrule
  \end{tabu}
  \caption{Convergence test in $M_{\text{sph}}$ of the 3D Maxwell molecule for $e=0.2$. Errors shown here are $|T_\text{num} - T_\text{ref}|$ at $t_\text{final}=1$. $\Delta t = 0.01$, $N = 32$, $N_{\rho}=30$, $R=7$, $L=7.72$.}
  \label{table 3}
\end{table}

\begin{table}[htp!]
  \centering
  \begin{tabu} to 0.8\linewidth {X[1,c] X[2,c]}
    \toprule
    $M_{\text{sph}}$ & $|Q_\text{num} - Q_\text{direct}|_{L^{\infty}}$\\
    \midrule
 	6 & 0.00041820767783433107 \\
	12 & 3.1726851245650724e-05 \\
	32 & 6.5752814213618921e-07 \\
	48 & 5.6132233266191489e-07 \\
	70 & 2.6714628356683257e-07 \\
	94 & 1.0508293902503248e-07 \\
	120 & 2.8872928147741922e-08 \\ 
    \bottomrule
  \end{tabu}
  \caption{Convergence test in $M_{\text{sph}}$ of the 3D Maxwell molecule for $e=0.2$. Errors shown here are $|Q_\text{num} - Q_\text{direct}|_{L^{\infty}}$ at $t_\text{final}=1$. $\Delta t = 0.01$, $N = 32$, $N_{\rho}=30$, $R=7$, $L=7.72$.}
  \label{table 4}
\end{table}

@@

\begin{figure}[htp!]
  \centering
  \includegraphics[width = .8\linewidth]{CPU_time}
  \caption{Same numbers as in Table~\ref{table 5} plotted in a log-log scale.}
  \label{CPUtime}
\end{figure}

@@

We take the same initial data \eqref{3Dbkw} as in the previous example but for hard sphere molecules, that is, consider the collision kernel $B_{\sigma} = \frac{1}{4\pi}|g|$. Also we set $\varepsilon=0$ in equation (\ref{inBoltz}) (the Haff's law is for unheated case, i.e., the homogeneous cooling state). We numerically evaluate the temperature up to time $t_\text{final}=3$ with two different values of restitution coefficient $e=0.2$ and $0.8$. The time evolution of $T$ is shown in Figure~\ref{Haff_cooling}, from which one can see that the numerical temperature decays as 
\begin{equation}
T(t) = \frac{T_0}{(1+C_0 t)^2},
\end{equation}
where the constant $C_0$ depends on the value of $e$. This is in perfect agreement of the Haff's formula (see Section 4.1 of \cite{NE}).

\begin{figure}[htp!]
  \centering
  \includegraphics[width = .8\linewidth]{Haff's_cooling}
  \caption{Haff's cooling law. $\Delta t=0.01$, $N=32$, $N_{\rho}=30$, $M_{\text{sph}}=32$, $R=7$, $L=7.72$.}
  \label{Haff_cooling}
\end{figure}

>>>>

\documentclass[review,times]{elsarticle}

@@

\usepackage[table]{xcolor}
\modulolinenumbers[5]

\usepackage{caption}

@@

Now assuming $f$ is a solution to equation \eqref{inBoltz}, the conservation of mass and momentum is straightforward by taking $\varphi = 1$ and $v$ respectively in the weak form \eqref{weak22}, i.e.,
\begin{equation}\label{conservation}
\rho \equiv \rho_0, \quad  m\equiv m_0,
\end{equation}
with $\rho_0$ and $m_0$ being the initial density and momentum. By taking $\varphi=|v|^2/2$ in \eqref{weak22}, one obtains the following energy equation
\begin{equation}\label{ki_energy}
\partial_t E -\varepsilon d\rho_0=-\frac{1}{16}\int_{\mathbb{R}^d}\int_{\mathbb{R}^d}\int_{S^{d-1}} (1-e^2)B_{\sigma}(|g|,\sigma\cdot \hat{g})|g|^2(1-\sigma\cdot \hat{g})ff_*\,\rd{\sigma}\rd{v}\rd{v_*},
\end{equation}
where we used (\ref{momentum}) and (\ref{relation}).

@@

\subsubsection{3D inelastic Maxwell molecules \color{blue} --- isotropic solution}

@@

\begin{table}[htp!]
  \centering
  \begin{tabu} to 0.8\linewidth {X[1,c] X[2,c]}
    \toprule
    $M_{\text{sph}}$ & $|T_\text{num} - T_\text{ref}|$\\
    \midrule
    6 & 0.0013957739849754791 \\
    12 & 9.9706271716293315e-05 \\
    32 & 2.2499901350947482e-06 \\
    48 & 2.4272557155313734e-06 \\
    70 & 2.4703481364962698e-06 \\
    94 & 2.4703481364962698e-06 \\
    120 & 2.453380453903975e-06 \\
    \bottomrule
  \end{tabu}
  \caption{Convergence test in $M_{\text{sph}}$ of the 3D Maxwell molecule {\color{blue} with isotropic initial} for $e=0.2$. Errors shown here are $|T_\text{num} - T_\text{ref}|$ at $t_\text{final}=1$. $\Delta t = 0.01$, $N = 32$, $N_{\rho}=30$, $R=7$, $L=7.72$.}
  \label{table 3}
\end{table}

\begin{table}[htp!]
  \centering
  \begin{tabu} to 0.8\linewidth {X[1,c] X[2,c]}
    \toprule
    $M_{\text{sph}}$ & $|Q_\text{num} - Q_\text{direct}|_{L^{\infty}}$\\
    \midrule
 	  6 & 0.00041820767783433107 \\
	  12 & 3.1726851245650724e-05 \\
	  32 & 6.5752814213618921e-07 \\
	  48 & 5.6132233266191489e-07 \\
	  70 & 2.6714628356683257e-07 \\
	  94 & 1.0508293902503248e-07 \\
	  120 & 2.8872928147741922e-08 \\ 
    \bottomrule
  \end{tabu}
  \caption{Convergence test in $M_{\text{sph}}$ of the 3D Maxwell molecule {\color{blue} with isotropic initial} for $e=0.2$. Errors shown here are $|Q_\text{num} - Q_\text{direct}|_{L^{\infty}}$ at $t_\text{final}=1$. $\Delta t = 0.01$, $N = 32$, $N_{\rho}=30$, $R=7$, $L=7.72$.}
  \label{table 4}
\end{table}

@@

\begin{figure}[htp!]
  \centering
  \includegraphics[width = .8\linewidth]{CPU_time}
  \caption{Same numbers as in Table~\ref{table 5} plotted in a log-log scale. {\color{blue} This test is performed using Python (numpy and scipy modules for array operations; pyfftw module (a Python wrapper of FFTW)for computing FFTs) on a $10$ CPUs workstation (Intel(R) Xeon(R) CPU E5-2660 v3 @ 2.60GHz, $8$ threads per FFTW operation.)}}
  \label{CPUtime}
\end{figure}

{\color{blue} \subsubsection{3D inelastic Maxwell molecules  --- anisotropic solution}

In previous subsection we perform a convergence test in $M_\text{sph}$ (compared with direct method and the analytical solution of the macroscopic quantity $T$) using an isotropic solution. This test shows us by using spherical design we can achieve very high accuracy of the integral on the 3D unit sphere surface, which is a key step in our algorithm for 3D case. In order to further investigate the performance of spherical design and to illustrate its necessity in the algorithm, an example with anisotropic initial is presented in this subsection.

For the sake of comparing with isotropic case, this test is done use exact the same configuration as in previous subsection except the initial condition:
\begin{equation}\label{3D_anisotropic}
  f_0(v) = \frac{4}{5\pi^{\frac{3}{2}}}\left(\exp(-\sqrt[3]{16}|v-2|^2) + \exp(-|v+0.5|^2)\right).
\end{equation}
We still set $e=0.2$, $t_\text{final} = 1$, $\Delta t = 0.01$, $N = 32$, and $N_{\rho}=30$ and first use the macroscopic quantity $T$ as our measurement. The results are shown in Table~\ref{3D_T_conv_anisotropic}. We can find both the convergence rate and accuracy are not as good as in isotropic case. It can only achieves an $O(10^{-4})$ error even using $192$ points of spherical design. One can also see this result from Table~\ref{3D_direct_conv_anisotropic} and Table~\ref{table 4}. Despite this result, we consider it as acceptable since the problem itself (numerical approximation of an integral on 3D unit sphere) is highly nontrivial, and we believe spherical design is so far the best method to deal with such problem. We illustrate this statement by performing another test using the standard gauss quadrature method which is also the method used in~\cite{WU201566} for the computation of the integral on 3D unit sphere, shown in Table~\ref{3D_quadrature_conv_anisotropic}. Under roughly the same number of quadrature points on sphere used \big((\# of $\theta$)$\times$(\# of $\phi$)\big), this quadrature method has a very bad performance: slow convergence and low accuracy.
}

\begin{table}[htpb]
  \color{blue}
  \centering
  \begin{tabu} to 0.8\linewidth {X[1,c] X[2,c]}
    \toprule
    $M_{\text{sph}}$ & $|T_\text{num} - T_\text{ref}|$\\
    \midrule
    6 & 2.661413377541729e-01 \\
    12 & 1.6773023124809094e-01 \\
    32 & 1.0297536867238799e-01 \\
    48 & 2.584202650005074e-02 \\
    70 & 2.422376071694199e-02 \\
    94 & 3.058799838901116e-03 \\
    120 & 4.413127388460447e-03 \\
    156 & 1.220110856786194e-03 \\
    192 & 3.96517639616345e-04 \\
    \bottomrule
  \end{tabu}
  \captionsetup{labelfont={color=blue}}
  \caption{\color{blue} Convergence test in $M_{\text{sph}}$ of the 3D Maxwell molecule with anisotropic initial for $e=0.2$. Errors shown here are $|T_\text{num} - T_\text{ref}|$ at $t_\text{final}=1$. $\Delta t = 0.01$, $N = 32$, $N_{\rho}=30$, $R=7$, $L=7.72$.}
  \label{3D_T_conv_anisotropic}
\end{table}

\begin{table}[htpb]
  \color{blue}
  \centering
  \begin{tabu} to 0.8\linewidth {X[1,c] X[2,c]}
    \toprule
    $M_{\text{sph}}$ & $|Q_\text{num} - Q_\text{direct}|_{L^{\infty}}$\\
    \midrule
 	  6 & 1.4674566108660258e-02 \\
    12 & 9.532325835757156e-03 \\
    32 & 5.92707599850977e-03 \\
    48 & 1.952601851310574e-03 \\
    70 & 1.43696494086e-03 \\
    94 & 4.354500695785647e-04 \\
    120 & 3.534377201557734e-04 \\
    156 & 9.811152052724729e-05 \\
    192 & 3.747408101774895e-05 \\ 
    \bottomrule
  \end{tabu}
  \captionsetup{labelfont={color=blue}}
  \caption{\color{blue} Convergence test in $M_{\text{sph}}$ of the 3D Maxwell molecule with anisotropic initial for $e=0.2$. Errors shown here are $|Q_\text{num} - Q_\text{direct}|_{L^{\infty}}$ at $t_\text{final}=1$. $\Delta t = 0.01$, $N = 32$, $N_{\rho}=30$, $R=7$, $L=7.72$.}
  \label{3D_direct_conv_anisotropic}
\end{table}

\begin{table}[htpb]
  \color{blue}
  \centering
  \begin{tabu} to 0.8\linewidth {X[1,c] X[2,c]}
    \toprule
    \# of $(\theta, \phi)$ & $|Q_\text{num} - Q_\text{direct}|_{L^{\infty}}$\\
    \midrule
    (2,3) & 9.301850859331023e-03 \\
    (3,4) & 1.4164542178285952e-02 \\
    (4,8) & 2.148725985732372e-02 \\
    (6,8) & 3.952940827407407e-03 \\
    (7,10) & 3.3540095954422665e-03 \\
    (10,10) & 3.405850043869026e-03 \\
    (10,12) & 3.016912442998702e-03 \\
    (11,15) & 1.227818352872178e-03 \\
    (12,16) & 1.2310868626689182e-03 \\
    \bottomrule
  \end{tabu}
  \captionsetup{labelfont={color=blue}}
  \caption{\color{blue} Convergence test of gauss quadrature method of the 3D Maxwell molecule with anisotropic solution for $e=0.2$. Errors shown here are $|Q_\text{num} - Q_\text{direct}|_{L^{\infty}}$ at $t_\text{final}=1$. $\Delta t = 0.01$, $N = 32$, $N_{\rho}=30$, $R=7$, $L=7.72$.}
  \label{3D_quadrature_conv_anisotropic}
\end{table}

@@

{\color{red} We perform two simulations regarding different initial data: The first one we take the same initial data \eqref{3Dbkw} as in the previous example; the second one we use the following initial data which is a Maxwellian with nonzero bulk velocity,
\begin{equation}\label{3D_maxwellian}
  f_0(v) = \frac{\rho_0}{(2\pi T_0)^{3/2}}e^{-(v - u_0)^2},
\end{equation}
where $\rho_0 = 1$, $T_0 = 2$ and $u_0 = (0.5, -0.5, 0)^T$. From~\eqref{conservation} we know $rho$ and $u$ with keep unchanged in time. Both simulations are performed} for hard sphere molecules, that is, consider the collision kernel $B_{\sigma} = \frac{1}{4\pi}|g|$. Also we set $\varepsilon=0$ in equation (\ref{inBoltz}) (the Haff's law is for unheated case, i.e., the homogeneous cooling state). We numerically evaluate the temperature up to time $t_\text{final}=3$ with two different values of restitution coefficient $e=0.2$ and $0.8$. The time evolution of $T$ is shown in Figure~\ref{Haff_cooling} and Figure~\ref{Haff_cooling_maxwellian}, from which one can see that the numerical temperature decays as 
\begin{equation}
T(t) = \frac{T_0}{(1+C_0 t)^2},
\end{equation}
where the constant $C_0$ {\color{red} depending on the value of $e$ is chosen by using least squares fitting}. This is in perfect agreement of the Haff's formula (see Section 4.1 of \cite{NE}).

\begin{figure}[htp!]
  \centering
  \includegraphics[width = .8\linewidth]{Haff's_cooling}
  \caption{Haff's cooling law {\color{red} with~\eqref{3Dbkw} as initial data}. $\Delta t=0.01$, $N=32$, $N_{\rho}=30$, $M_{\text{sph}}=32$, $R=7$, $L=7.72$.}
  \label{Haff_cooling}
\end{figure}

\begin{figure}[htp!]
  \centering
  \includegraphics[width = .8\linewidth]{Haff_cooling_maxwellian}
  \captionsetup{labelfont={color=blue}}
  \caption{\color{blue} Haff's cooling law {\color{red} with~\eqref{3D_maxwellian} as initial data}. $\Delta t=0.01$, $N=32$, $N_{\rho}=30$, $M_{\text{sph}}=32$, $R=7$, $L=7.72$.}
  \label{Haff_cooling_maxwellian}
\end{figure}
